---
title: "IDS 572 Assignment - 2"
author: "Nipun Ranjan, Amit Patel, Tejashvy Raj Singh"
date: "2022-09-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Q1 
Importing libraries
```{r}
library(tidyverse)
library(rpart)
library(rpart.plot)
```
Importing datasets
```{r}
train_bc_data <- read.csv("01 Case -1 Train data/trainX.csv")
train_bc_label <- read.csv("01 Case -1 Train data/trainY.csv")

colnames(train_bc_data) <- c("Radius.mean","Texture.mean","Perimeter.mean","Area.mean","Smoothness.mean","Compactness.mean","Concavity.mean","nocp.mean","Symmetry.mean","fd.mean","Radius.sd","Texture.sd","Perimeter.sd","Area.sd","Smoothness.sd","Compactness.sd","Concavity.sd","nocp.sd","Symmetry.sd","fd.sd","Radius.max","Texture.max","Perimeter.max","Area.max","Smoothness.max","Compactness.max","Concavity.max","nocp.max","Symmetry.max","fd.max")

colnames(train_bc_label) <- c("Label")

test_bc_data <- read.csv("02 Case -1 Test data/testX.csv")
test_bc_label <- read.csv("02 Case -1 Test data/testY.csv")

colnames(test_bc_data) <- c("Radius.mean","Texture.mean","Perimeter.mean","Area.mean","Smoothness.mean","Compactness.mean","Concavity.mean","nocp.mean","Symmetry.mean","fd.mean","Radius.sd","Texture.sd","Perimeter.sd","Area.sd","Smoothness.sd","Compactness.sd","Concavity.sd","nocp.sd","Symmetry.sd","fd.sd","Radius.max","Texture.max","Perimeter.max","Area.max","Smoothness.max","Compactness.max","Concavity.max","nocp.max","Symmetry.max","fd.max")

colnames(test_bc_label) <- c("Label")

train_bc <- cbind(train_bc_data,train_bc_label)
```
Data exploration
```{r}
train_bc_fivnums <- sapply(train_bc[c("Radius.mean","Texture.mean","Perimeter.mean","Area.mean","Smoothness.mean","Compactness.mean","Concavity.mean","nocp.mean","Symmetry.mean","fd.mean","Radius.sd","Texture.sd","Perimeter.sd","Area.sd","Smoothness.sd","Compactness.sd","Concavity.sd","nocp.sd","Symmetry.sd","fd.sd","Radius.max","Texture.max","Perimeter.max","Area.max","Smoothness.max","Compactness.max","Concavity.max","nocp.max","Symmetry.max","fd.max")],fivenum)%>%
  view()

Distance_q1_q3 <- train_bc_fivnums[4,]-train_bc_fivnums[2,]


train_bc_fivnums <- rbind(train_bc_fivnums,Distance_q1_q3)
view(train_bc_fivnums)

IQR <- 1.5*train_bc_fivnums[6,]
train_bc_fivnums <- rbind(train_bc_fivnums,IQR)%>%
  view()

dist_q3_max <- train_bc_fivnums[5,]-train_bc_fivnums[4,]
dist_q1_min <- train_bc_fivnums[2,]-train_bc_fivnums[1,]

train_bc_fivnums <- rbind(train_bc_fivnums,dist_q3_max)
train_bc_fivnums <- rbind(train_bc_fivnums,dist_q1_min)

view(train_bc_fivnums)

rownames(train_bc_fivnums) <- c("Min","Quartile-1","Median","Quartile-3","Max","diff_q1_q3","Outlier_dist","dist_q3_max","dist_q1_min")

view(train_bc_fivnums)
```
By looking at the table created in the above step, we can conclusively see that there are features which have outliers using the IQR rule. The next step will be mark on the actual dataset these outliers
```{r}
o1<-which(train_bc$Radius.mean %in% boxplot.stats(train_bc$Radius.mean)$out)
o2<-which(train_bc$Texture.mean %in% boxplot.stats(train_bc$Texture.mean)$out)
o3<-which(train_bc$Perimeter.mean %in% boxplot.stats(train_bc$Perimeter.mean)$out)
o4<-which(train_bc$Area.mean %in% boxplot.stats(train_bc$Area.mean)$out)
o5<-which(train_bc$Smoothness.mean %in% boxplot.stats(train_bc$Smoothness.mean)$out)
o6<-which(train_bc$Concavity.mean %in% boxplot.stats(train_bc$Concavity.mean)$out)
o7<-which(train_bc$nocp.mean %in% boxplot.stats(train_bc$nocp.mean)$out)
o8<-which(train_bc$Symmetry.mean %in% boxplot.stats(train_bc$Symmetry.mean)$out)
o9<-which(train_bc$fd.mean %in% boxplot.stats(train_bc$fd.mean)$out)
o10<-which(train_bc$Compactness.mean %in% boxplot.stats(train_bc$Compactness.mean)$out)

o11<-which(train_bc$Radius.sd %in% boxplot.stats(train_bc$Radius.sd)$out)
o12<-which(train_bc$Texture.sd %in% boxplot.stats(train_bc$Texture.sd)$out)
o13<-which(train_bc$Perimeter.sd %in% boxplot.stats(train_bc$Perimeter.sd)$out)
o14<-which(train_bc$Area.sd %in% boxplot.stats(train_bc$Area.sd)$out)
o15<-which(train_bc$Smoothness.sd %in% boxplot.stats(train_bc$Smoothness.sd)$out)
o16<-which(train_bc$Concavity.sd %in% boxplot.stats(train_bc$Concavity.sd)$out)
o17<-which(train_bc$nocp.sd %in% boxplot.stats(train_bc$nocp.sd)$out)
o18<-which(train_bc$Symmetry.sd %in% boxplot.stats(train_bc$Symmetry.sd)$out)
o19<-which(train_bc$fd.sd %in% boxplot.stats(train_bc$fd.sd)$out)
o20<-which(train_bc$Compactness.sd %in% boxplot.stats(train_bc$Compactness.sd)$out)

o21<-which(train_bc$Radius.max %in% boxplot.stats(train_bc$Radius.max)$out)
o22<-which(train_bc$Texture.max %in% boxplot.stats(train_bc$Texture.max)$out)
o23<-which(train_bc$Perimeter.max %in% boxplot.stats(train_bc$Perimeter.max)$out)
o24<-which(train_bc$Area.max %in% boxplot.stats(train_bc$Area.max)$out)
o25<-which(train_bc$Smoothness.max %in% boxplot.stats(train_bc$Smoothness.max)$out)
o26<-which(train_bc$Concavity.max %in% boxplot.stats(train_bc$Concavity.max)$out)
o27<-which(train_bc$nocp.max %in% boxplot.stats(train_bc$nocp.max)$out)
o28<-which(train_bc$Symmetry.max %in% boxplot.stats(train_bc$Symmetry.max)$out)
o29<-which(train_bc$fd.max %in% boxplot.stats(train_bc$fd.max)$out)
o30<-which(train_bc$Compactness.max %in% boxplot.stats(train_bc$Compactness.max)$out)

view(train_bc)

train_bc_cleaned <- train_bc[-c(o1,o2,o3,o4,o5,o6,o7,o8,o9,o10,o11,o12,o13,o14,o15,o16,o17,o18,o19,o20,o21,o22,o23,o24,o25,o26,o27,o28,o29,o30),]

view(train_bc_cleaned)
```

```{r}
train_bc_navals <- sapply(train_bc[c("Radius.mean","Texture.mean","Perimeter.mean","Area.mean","Smoothness.mean","Compactness.mean","Concavity.mean","nocp.mean","Symmetry.mean","fd.mean","Radius.sd","Texture.sd","Perimeter.sd","Area.sd","Smoothness.sd","Compactness.sd","Concavity.sd","nocp.sd","Symmetry.sd","fd.sd","Radius.max","Texture.max","Perimeter.max","Area.max","Smoothness.max","Compactness.max","Concavity.max","nocp.max","Symmetry.max","fd.max")],is.na)%>%
  view()

train_bc_navals <- sapply(train_bc_navals[c("Radius.mean","Texture.mean","Perimeter.mean","Area.mean","Smoothness.mean","Compactness.mean","Concavity.mean","nocp.mean","Symmetry.mean","fd.mean","Radius.sd","Texture.sd","Perimeter.sd","Area.sd","Smoothness.sd","Compactness.sd","Concavity.sd","nocp.sd","Symmetry.sd","fd.sd","Radius.max","Texture.max","Perimeter.max","Area.max","Smoothness.max","Compactness.max","Concavity.max","nocp.max","Symmetry.max","fd.max")],unique)%>%
  view()

```
There are no NA values in the dataset
```{r}
correlation_coefficient_train_bc <- cor(train_bc)%>%
  view()

correlation_coefficient_train_bc_cleaned <- cor(train_bc_cleaned)%>%
  view()

correlation_coff_com <- rbind(correlation_coefficient_train_bc[31,],correlation_coefficient_train_bc_cleaned[31,])%>%
  view()

```
Extracting features with the highest correlation amongst the three cohorts.
```{r}
tree_1_org_data <- rpart(as.factor(Label)~Radius.mean+Perimeter.mean+Area.mean+Concavity.mean+nocp.mean+Radius.sd+Perimeter.sd+Area.sd+Compactness.sd+Radius.max+Perimeter.max+Area.max+Concavity.max+nocp.max+Compactness.max, data = train_bc, parms =list(split = "information"), control = rpart.control(minbucket = 3, minsplit = 8))

rpart.plot(tree_1_org_data)

prediction_1_test <- predict(tree_1_org_data,test_bc_data, type="class")
prediction_2_train <- predict(tree_1_org_data,train_bc_data, type="class")


view(prediction_1_test)

view(test_bc_label)

confs_mat_1 <- table(prediction_1_test,test_bc_label$Label)%>%
  view()

confs_mat_1a <- table(prediction_2_train,train_bc_label$Label)%>%
  view()

Accuracy_tree_1_test <- 53/56
Accuracy_tree_1_train <- 443/454
print(Accuracy_tree_1_train)
print(Accuracy_tree_1_test)
```

```{r}
tree_2_clean_data <-rpart(as.factor(Label)~Radius.mean+Perimeter.mean+Area.mean+Concavity.mean+nocp.mean+Radius.sd+Perimeter.sd+Area.sd+Compactness.sd+Radius.max+Perimeter.max+Area.max+Concavity.max+nocp.max+Compactness.max, data = train_bc_cleaned, parms =list(split = "information"),control=rpart.control(minbucket = 3, minsplit = 7))

rpart.plot(tree_2_clean_data)

prediction_2_test <- predict(tree_2_clean_data,test_bc_data, type="class")
prediction_2_train <- predict(tree_2_clean_data,train_bc_cleaned$Label,type="class")

view(prediction_2_test)

confs_mat_2 <- table(prediction_2_test,test_bc_label$Label)%>%
  view()

Accuracy_2_tree <- 51/56
print(Accuracy_2_tree)
```


```{r}
tree_3_org_data <- rpart(as.factor(Label)~.,data = train_bc,control = rpart.control(minbucket = 3, minsplit = 7))
rpart.plot(tree_3_org_data)

prediction_3_test <- predict(tree_3_org_data,test_bc_data, type="class")
prediction_3_train <- predict(tree_3_org_data,train_bc_data, type="class")


view(prediction_3_test)

confs_mat_3 <- table(prediction_3_test,test_bc_label$Label)%>%
  view()

Accuracy_3_tree <- 50/56
print(Accuracy_3_tree)



```

```{r}
tree_4_clean_data <- rpart(as.factor(Label)~.,data = train_bc_cleaned,control = rpart.control(minbucket = 3, minsplit = 7))
rpart.plot(tree_4_clean_data)

prediction_4_test <- predict(tree_4_clean_data,test_bc_data, type="class")
prediction_4_train <- predict(tree_4_clean_data,train_bc_data, type="class")


view(prediction_4_test)

confs_mat_4 <- table(prediction_4_test,test_bc_label$Label)%>%
  view()

Accuracy_2_tree <- 47/56
print(Accuracy_2_tree)



```
Tree_1 gave the best accuracy when tested with test data, hence will go ahead with that

```{r}
view(tree_1_org_data$cptable)
tree_1_org_data_pruned <- prune(tree_1_org_data,cp=0.0100000)
rpart.plot(tree_1_org_data_pruned)

prediction_5_test <- predict(tree_1_org_data_pruned,test_bc_data, type="class")
prediction_5_train <- predict(tree_1_org_data_pruned,train_bc_data, type="class")


view(prediction_5_test)

confs_mat_5 <- table(prediction_5_test,test_bc_label$Label)%>%
  view()

confs_mat_5a <- table(prediction_5_train,train_bc_label$Label)%>%
  view()

Accuracy_5_tree <- 53/56
Accuracy_5_train <- 443/454
print(Accuracy_5_train)
print(Accuracy_5_tree)
```


Question - 1

a. There were no NA values in the dataset. We did identify outliers in all the variable, and have produced trees after removing these from the dataset; however, this caused the accuracy of the tree to go down on the test dataset. The highly correlated variables with the Label are chosen to produce the most optimal tree.

b. The tree has 5 leaf nodes.
c. Major predictors :
    i. Max value of Perimeter
    ii. Max value of 'Number of concave portions'
    iii. Max value of Area
    iv. Mean value of Radius
    
  These are from the subset of highly correlated variables with the Label.

d. Two rules for people likely to have cancer:
    i. if Perimeter.max >106 and NOCP.max >0.15 then Cancer
    ii.if Perimeter.max >106 and NOCP.max <0.15 and Area.max>957 then Cancer
    
e. Accuracy of the tree :
    i. Training dataset: 
    ii.Test dataset : 92.85%
  
f. To create the best possible tree, we looked at the best CP value for the        tree with the highest accuracy. The best CP value was chosen based on the      minimum error. We pruned the tree based on this value. To evaluate its         performance, we looked at the accuracy of its predictions on the test          dataset.

g. All rules :
    i. if perimeter.max <106 then No cancer
    ii.if perimeter.max=>106, nocp.max<0.15, area.max<957, Radius.mean=>14             then No cancer
    iii.if perimeter.max=>106, nocp.max<0.15, area.max<957, Radius.mean<14             then Cancer
    iv. if Perimeter.max >106 and NOCP.max >0.15 then Cancer
    v.if Perimeter.max >106 and NOCP.max <0.15 and Area.max>957 then Cancer


###############################################################################



Q2
Importing dataset
```{r}
zoo_full_file_t3 <- read.csv("03 Case -2 Train data/zoo.csv")
zoo_train_file_t1 <- read.csv("03 Case -2 Train data/zoo1.csv")
zoo_test_file_t2 <- read.csv("04 Case -2 Test data/zoo2.csv")
```
Converting variables to factors
```{r}
zoo_train_file_t1 <- mutate(zoo_train_file_t1,hair=as.factor(hair))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,feathers=as.factor(feathers))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,eggs=as.factor(eggs))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,milk=as.factor(milk))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,airborne=as.factor(airborne))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,aquatic=as.factor(aquatic))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,predator=as.factor(predator))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,toothed=as.factor(toothed))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,backbone=as.factor(backbone))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,breathes=as.factor(breathes))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,venomous=as.factor(venomous))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,legs=as.numeric(legs))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,tail=as.factor(tail))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,catsize=as.factor(catsize))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,domestic=as.factor(domestic))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,fins=as.factor(fins))
zoo_train_file_t1 <- mutate(zoo_train_file_t1,type=as.factor(type))

zoo_train_file_t1$animal <- NULL
```

```{r}
zoo_test_file_t2 <- mutate(zoo_test_file_t2,hair=as.factor(hair))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,feathers=as.factor(feathers))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,eggs=as.factor(eggs))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,milk=as.factor(milk))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,airborne=as.factor(airborne))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,aquatic=as.factor(aquatic))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,predator=as.factor(predator))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,toothed=as.factor(toothed))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,backbone=as.factor(backbone))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,breathes=as.factor(breathes))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,venomous=as.factor(venomous))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,legs=as.numeric(legs))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,tail=as.factor(tail))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,catsize=as.factor(catsize))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,domestic=as.factor(domestic))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,fins=as.factor(fins))
zoo_test_file_t2 <- mutate(zoo_test_file_t2,type=as.factor(type))

```

```{r}
zoo_full_file_t3 <- mutate(zoo_full_file_t3,hair=as.factor(hair))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,feathers=as.factor(feathers))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,eggs=as.factor(eggs))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,milk=as.factor(milk))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,airborne=as.factor(airborne))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,aquatic=as.factor(aquatic))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,predator=as.factor(predator))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,toothed=as.factor(toothed))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,backbone=as.factor(backbone))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,breathes=as.factor(breathes))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,venomous=as.factor(venomous))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,legs=as.numeric(legs))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,tail=as.factor(tail))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,catsize=as.factor(catsize))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,domestic=as.factor(domestic))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,fins=as.factor(fins))
zoo_full_file_t3 <- mutate(zoo_full_file_t3,type=as.factor(type))

```

```{r}
zoo_tree_type <- rpart(as.factor(type)~.,zoo_train_file_t1, method='class',parms = list(split = "information"),control = rpart.control(minbucket = 2, minsplit = 8))
rpart.plot(zoo_tree_type)

prediction_1_zoo_test <- predict(zoo_tree_type,zoo_test_file_t2, type="class")
prediction_1_zoo_train <- predict(zoo_tree_type,zoo_train_file_t1,type="class")


view(prediction_1_zoo_test)
view(zoo_test_file_t2)

prediction_comparison <- cbind(prediction_1_zoo_test, zoo_test_file_t2$type)%>%
  view()

prediction_comparison_train <- cbind(prediction_1_zoo_train,zoo_train_file_t1$type)%>%
  view()


#write.csv(prediction_1_zoo_test,"04 Case -2 Test #data/test_results_zoo_test.csv")

Accuracy_1_zoo_tree <- 26/30
Accuracy_1_zoo_train <- 68/71
print(Accuracy_1_zoo_train)
print(Accuracy_1_zoo_tree)

```

```{r}

view(zoo_tree_type$cptable)
zoo_tree_type_pruned <- prune(zoo_tree_type, cp = 0.0100000 )
rpart.plot(zoo_tree_type_pruned)

prediction_1_zoo_test_pruned <- predict(zoo_tree_type_pruned,zoo_test_file_t2, type="class")
prediction_1_zoo_train_pruned <- predict(zoo_tree_type_pruned,zoo_train_file_t1,type="class")



prediction_comparison_pruned <- cbind(prediction_1_zoo_test_pruned, zoo_test_file_t2$type)%>%
  view()

prediction_comparison_train_pruned <- cbind(prediction_1_zoo_train_pruned,zoo_train_file_t1$type)%>%
  view()


#write.csv(prediction_1_zoo_test,"04 Case -2 Test #data/test_results_zoo_test.csv")

Accuracy_1_zoo_tree_pruned <- 26/30
Accuracy_1_zoo_train_pruned <- 68/71
print(Accuracy_1_zoo_train_pruned)
print(Accuracy_1_zoo_tree_pruned)
```
No changes in the tree

1R tree - Milk
```{r}

zoo_tree_type_milk_1R <- rpart(as.factor(type)~milk,zoo_full_file_t3, method='class',parms = list(split = "information"),control = rpart.control(minbucket = 2, minsplit = 8))
rpart.plot(zoo_tree_type_milk_1R)

prediction_1_zoo_train_milk <- predict(zoo_tree_type_milk_1R,zoo_full_file_t3,type="class")


prediction_comparison_train_milk <- cbind(prediction_1_zoo_train_milk,zoo_full_file_t3$type)%>%
  view()


Accuracy_1_zoo_tree_milk <- 17/30
Accuracy_1_zoo_train_milk <- 35/71
print(Accuracy_1_zoo_train_milk)
print(Accuracy_1_zoo_tree_milk)

```
1R - Feather
```{r}

zoo_tree_type_feather_1R <- rpart(as.factor(type)~feather,zoo_full_file_t3, method='class',parms = list(split = "information"),control = rpart.control(minbucket = 2, minsplit = 8))
rpart.plot(zoo_tree_type_feather_1R)

prediction_1_zoo_train_feather <- predict(zoo_tree_type_feather_1R,zoo_full_file_t3,type="class")




prediction_comparison_train_feather <- cbind(prediction_1_zoo_train_feather,zoo_full_file_t3$type)%>%
  view()


Accuracy_1_zoo_tree_feather <- 17/30
Accuracy_1_zoo_train_feather <- 35/71
print(Accuracy_1_zoo_train_feather)
print(Accuracy_1_zoo_tree_feather)

```




1R- Aquatic
```{r}
zoo_tree_type_aquatic_1R <- rpart(as.factor(type)~aquatic,zoo_full_file_t3, method='class',parms = list(split = "information"),control = rpart.control(minbucket = 2, minsplit = 8))
rpart.plot(zoo_tree_type_aquatic_1R)

prediction_1_zoo_train_aquatic <- predict(zoo_tree_type_aquatic_1R,zoo_full_file_t3,type="class")


prediction_comparison_train_aquatic <- cbind(prediction_1_zoo_train_aquatic,zoo_full_file_t3$type)%>%
  view()


Accuracy_1_zoo_tree_aquatic <- 17/30
Accuracy_1_zoo_train_aquatic <- 35/71
print(Accuracy_1_zoo_train_aquatic)
print(Accuracy_1_zoo_tree_aquatic)

```



b. 
  i. Leaves in the tree : 4
  ii.Classification accuracy on training dataset : 
  iii.Classification accuracy on test dataset : 80%

d. 

  
